A Flux Scale for Southern Hemisphere 21cm EoR Experiments

### 2013-06-13 ms.tex
Created a new "Background" section that incorporates and organizes S1.1, S1.2, and some of S2.  Re-wrote beginning of S2 to give an overview of the approach.

\subsection{The sources and instrument}
EoR measurements in the southern sky have focused on the coldest regions where
galactic foregrounds are minimal. With the majority of possible observing time falling around
RA=4h,Dec=-30. The brightest and least-resolved calibrator in this region is Pictor A 
(5h19m49.1,-45d46m45.0). Pictor A is a nearby FR-II type radio galaxy 
 similar to Cygnus A.  At $\sim$400Jy, Pictor is bright and sufficiently distant from other 
bright sources to make it eminently suitable as both a phase and flux calibrator. Its apparent
size of $\sim$8\arcmin{} is smaller than the scales being probed by current EoR instruments, 
making it suitable for precision calibration, with only a modest
level of resolution effects.
However, 
like most other sources, precise flux measurements in the EoR band are not available.
The previous best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.

@@

Establishing an accurate spectrum for Pictor A is of particular importance for
PAPER --- a dedicated EoR experiment 
that employs drift-scanning, dual-polarization dipole antennas 
tuned for efficient operation over a 120--170-MHz band.  PAPER is located in the South African Karoo desert
on the Square Kilometer Array
South Africa (SKA-SA) reserve, 100km north of the small town of Carnarvon.
The PAPER array has grown from 16 elements deployed in early 2009 to a
64-element imaging array in 2011 (see Figure \ref{fig:antpos}). 
Since November 2011 it has been arranged in a maximally redundant grid 
configuration to make deep power spectral integrations \citep{Parsons:2012p9028}.
Though highly sensitive as a power spectrum instrument, the maximally redundant array 
has a broad point spread function in the image domain. This severely limits the number 
of sources which may be used for flux calibration. Drift scanning across the sky with a 45\arcdeg{} FWHM
primary beam, there are very few unresolved, bright sources which are far from the galactic plane. 
Pictor A is bright and well enough separated from other emission to dominate the visibilities for a good fraction of
the EoR observing season, making it an excellent flux calibrator.

\subsection{Primary beam}
\label{sec:background}

Historically, the best Southern Hemisphere 
EoR band data were by \citet{Slee:1995p7541} with Culgoora Circular
Array\footnote{Known during daylight hours as Culgoora Radio Heliograph} and
various higher frequency measurements with Parkes.  These data are typically
uncertain to 20\% or higher and provide little coverage of the EoR band beyond
a single narrow-band data point.

More recent surveys include narrow-band surveys by the GMRT \footnote{\url{http://tgss.ncra.tifr.res.in/}} and Mauritius \citep{Pandey:2005p8687}, a
deep survey of the region near Hydra A by the 32 antenna MWA prototype
\cite{Williams:2012p8768} and a wide field survey by PAPER, also with 32
elements \cite{Jacobs:2011p8438}. Several sub-channels were provided in the
Williams catalog, though with 60-80\% error bars ---large compared to the 30\%
uncertainty on their wide band measurements.  The latter cover the band and
spatial scales relevant to EoR measurements but are limited by the accuracy of
the primary beam \citep{Jacobs:2013p9908} as well as the lack of precise in-band flux calibrators.

The response of the primary beam is of critical importance to EoR measurements.
Differences between the polarization responses cause leakage of polarized
signals into the total intensity measurement possibly corrupting the 
EoR power spectrum \citep{Moore:2013p9621}.  The primary beam 
shape is also critical to measuring and subtracting foregrounds 
\cite{Bernardi:2012p9020,Sullivan:2012p9457,Morales:2012p8790} a process which, to be effective, must be done to better than 1\%
precision for the brightest sources \citet{Liu:2009p4762,Bowman:2009p7816} .

A method for decoupling uncertain fluxes from the uncertain beam has been
described by \citet{Pober:2012p8800}. In simulation the method was able to
achieve 3 to 10\% accuracy in measuring the primary beam, depending on the
number of antennae and other variables. The Pober method emphasized the need
% XXX mention key assumption of 180deg rotational symmetry
for many repeated measurements of each alt-az pointing. Further investigation
is under way to improve and implement this method and would be greatly aided by
the availability of precise flux measurements unaffected by primary beam
uncertainty.

In this paper we provide a set of detailed total flux measurements aimed at providing
an absolute flux calibration for southern hemisphere EoR studies, as well as a
small set of verification fluxes suitable for constraining future primary beam studies.

\section{Approach}
\label{sec:approach}
In this section, we describe how we use PAPER in its imaging configuration
to measure Pictor A and a  selection of known, bright sources. The fluxes are obtained
by forming beams towards known sources as they drift through the primary beam, calibrating to a single source making a similar track.
These long tracks allow us to sidestep primary beam uncertainty and measure accurately calibrated fluxes.

 In \citet{Jacobs:2013p9908} several errors were identified as originating in primary beam uncertainty.
  First, under most imaging schemes, each source flux is measured at just a few 
points in the primary beam. As errors tend to vary across the beam it is often
 difficult to decouple flux uncertainty from beam uncertainty.
 Using the drift scan-beamforming technique, each 
source is measured thousands of times at a variety of different primay beam values to provide a complete sample of the
flux variance due to primary beam variation.  Second, the flux calibration was found to vary over the 
sky due to the dual uncertainties of primary beam response and prior catalog. To reduce our exposure to flux calibration
variation, we limit our observations to sources passing within 5\arcdeg{}  of pointings which we can directly calibrate
to a single reference source.
 A third limitation, identified in \citet{Williams:2012p8768},
was the increase of uncertainty towards the edge of the beam. To minimize our sensitivity to both beam and noise
uncertainty, we weight the source track by an additional factor of the primary beam model.

\subsection{Selection of confirmation sources}
As PAPER is a drift scan instrument, each declination describes a distinct path through the primary beam. The flux time seres of a beamform
 provides a detailed sample of the primary beam relative to the peak of the trace, however,
a model is still required to calibrate between declinations. Below we argue that be averaging over long tracks, we minimize our 
susceptibility to localized primary beam error and find that limiting our source selection range to within 
a declination range of 5\arcdeg{} around a calibrator is a good compromise. In section\ref{fig:gain}
we estimate the error resulting from this limited scope application of the primary beam model.

@@

Here we have provided a measurement of Pictor A with enough precision to confirm
a highly linear spectral index between 150 and 600MHz. We then used the same
filtered beam form method to measure the spectra of bright sources with similar
primary beam response.

>>>>

\section{Background}
\label{sec:background}

Historically, the best Southern Hemisphere 
EoR band data were by \citet{Slee:1995p7541} with Culgoora Circular
Array\footnote{Known during daylight hours as Culgoora Radio Heliograph} and
various higher frequency measurements with Parkes.  These data are typically
uncertain to 20\% or higher and provide little coverage of the EoR band beyond
a single narrow-band data point. 
More recent surveys include narrow-band surveys by the GMRT \footnote{\url{http://tgss.ncra.tifr.res.in/}} and Mauritius \citep{Pandey:2005p8687}, a
deep survey of the region near Hydra A by the 32 antenna MWA prototype
\cite{Williams:2012p8768} and a wide field survey by PAPER, also with 32
elements \cite{Jacobs:2011p8438}. Several sub-channels were provided in the
Williams catalog, though with 60-80\% error bars ---large compared to the 30\%
uncertainty on their wide band measurements.  The latter cover the band and
spatial scales relevant to EoR measurements but are limited by the accuracy of
the primary beam \citep{Jacobs:2013p9908} as well as the lack of precise in-band flux calibrators.

The response of the primary beam is of critical importance to EoR measurements.
Differences between the polarization responses cause leakage of polarized
signals into the total intensity measurement possibly corrupting the 
EoR power spectrum \citep{Moore:2013p9621}.  The primary beam 
shape is also critical to measuring and subtracting foregrounds 
\cite{Bernardi:2012p9020,Sullivan:2012p9457,Morales:2012p8790} a process which, to be effective, must be done to better than 1\%
precision for the brightest sources \citet{Liu:2009p4762,Bowman:2009p7816} .
A method for decoupling uncertain fluxes from the uncertain beam has been
described by \citet{Pober:2012p8800}. In simulation the method was able to
achieve 3 to 10\% accuracy in measuring the primary beam, depending on the
number of antennae and other variables. The Pober method emphasized the need
% XXX mention key assumption of 180deg rotational symmetry
for many repeated measurements of each alt-az pointing. Further investigation
is under way to improve and implement this method and would be greatly aided by
the availability of precise flux measurements unaffected by primary beam
uncertainty.

EoR measurements by PAPER and the MWA in the southern hemisphere have focused on the coldest regions where
galactic foregrounds are minimal. With the majority of possible observing time falling around
RA=4h,Dec=-30. The brightest and least-resolved calibrator in this region is Pictor A 
(5h19m49.1,-45d46m45.0). Pictor A is a nearby FR-II type radio galaxy 
 similar to Cygnus A.  At $\sim$400Jy, Pictor is bright and sufficiently distant from other 
bright sources to make it eminently suitable as both a phase and flux calibrator. Its apparent
size of $\sim$8\arcmin{} is smaller than the scales being probed by current EoR instruments, 
making it suitable for precision calibration, with only a modest
level of resolution effects.
However, 
like most other sources, precise flux measurements in the EoR band are not available.
The previous best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.

@@

Establishing an accurate spectrum for Pictor A is of particular importance for
PAPER --- a dedicated EoR experiment 
that employs drift-scanning, dual-polarization dipole antennas 
tuned for efficient operation over a 120--170-MHz band.  PAPER is located in the South African Karoo desert
on the Square Kilometer Array
South Africa (SKA-SA) reserve, 100km north of the small town of Carnarvon.
The PAPER array has grown from 16 elements deployed in early 2009 to a
64-element imaging array in 2011 (see Figure \ref{fig:antpos}). 
Since November 2011 it has been arranged in a maximally redundant grid 
configuration to make deep power spectral integrations \citep{Parsons:2012p9028}.
Though highly sensitive as a power spectrum instrument, the maximally redundant array 
has a broad point spread function in the image domain. This severely limits the number 
of sources which may be used for flux calibration. Drift scanning across the sky with a 45\arcdeg{} FWHM
primary beam, there are very few unresolved, bright sources which are far from the galactic plane. 
Pictor A is bright and well enough separated from other emission to dominate the visibilities for a good fraction of
the EoR observing season, making it a desireable source to use for flux calibration.

\section{Approach}
\label{sec:approach}
In this section, we describe our general approach to controlling the impact of beam model errors on measured
spectra derived from 
drift-scan observations 
of Pictor A and a selection of known, bright sources.  Our approach uses a set of ``source tracks'' as
a function of frequency and time.  Each source track is a beam, formed by phasing measured visibilities toward
known source locations as they drift through the primary beam, and summing over antenna pairs, as
described in \S\ref{sec:XXX}.  Since sources fall
at different positions
within the primary beam, it is generally not possible to relate the fluxes of sources at different positions to
one another without an accurate beam model.  To date, the accuracy of source flux measurements in the
southern hemisphere in the 100--200 MHz band have largely been limited by the accuracy of these beam models.

We mitigate this problem by selecting sources within a narrow declination range, so that source tracks represent
nearly identical cuts through the beam response pattern.  Using this fact, the relative amplitudes of sources
can be deduced with minimal reliance on an accurate beam model.  In our analysis, a prior beam model is only
necessary for extrapolating across a narrow declination range (we estimate the errors associated with this 
extrapolation in \S\ref{sec:XXX}), and for approximating optimal inverse-variance weighting when averaging
over source tracks to determine a single spectrum.

This approach addresses several sources of error that were identified in \citet{Jacobs:2013p9908} 
as originating from uncertainty in the primary beam model.
  First, under most imaging schemes, each source flux is measured at just a few 
points in the primary beam. As errors tend to vary across the beam it is often
 difficult to decouple flux uncertainty from beam uncertainty.
 Using the drift scan-beamforming technique, each 
source is measured thousands of times at a variety of different primay beam values to provide a complete sample of the
flux variance due to primary beam variation.  Second, the flux calibration was found to vary over the 
sky due to the dual uncertainties of primary beam response and prior catalog. To reduce our exposure to flux calibration
variation, we limit our observations to sources passing within 5\arcdeg{}  of pointings which we can directly calibrate
to a single reference source.
 A third limitation, identified in \citet{Williams:2012p8768},
was the increase of uncertainty towards the edge of the beam. To minimize our sensitivity to both beam and noise
uncertainty, we weight the source track by an additional factor of the primary beam model.

The remainder of the analysis in this paper relates to calibrating the relative source amplitudes that we
establish to an absolute flux scale as a function of frequency.  Given the shortcomings of 
existing in-band measurements of sources
in the southern hemisphere, we must bootstrap this flux scale from an ensemble of calibrator sources that
exhibit simple power-law spectra.
As described in \S\ref{sec:XXX},
this proceeds by first using a single calibrator to set the relative amplitudes of spectral channels, and
then using several sources, extrapolating across a large set of measurements 
between 70 MHz and 2 GHz, in a Markov-Chain Monte Carlo simulation to establish an absolute flux scale
for the measured spectra.

\subsection{Selection of confirmation sources} % XXX what is meant by "confirmation source" ?  Maybe just "Source Selection"?
As PAPER is a drift scan instrument, each declination describes a distinct path through the primary beam. The flux time seres of a beamform
 provides a detailed sample of the primary beam relative to the peak of the trace, however,
a model is still required to calibrate between declinations. Below we argue that be averaging over long tracks, we minimize our 
susceptibility to localized primary beam error and find that limiting our source selection range to within 
a declination range of 5\arcdeg{} around a calibrator is a good compromise. In \S\ref{fig:gain},
we estimate the error resulting from this limited scope application of the primary beam model.

@@

This paper presents a set of detailed total flux measurements, providing
an absolute flux calibration for southern hemisphere EoR studies, as well as a
modest set of verification fluxes for other sources suitable for constraining future primary beam studies.
Here we have provided a measurement of Pictor A with enough precision to confirm
a highly linear spectral index between 150 and 600MHz. We then used the same
filtered beam form method to measure the spectra of bright sources with similar
primary beam response.
Although we focus on a narrow declination range in this paper, the techniques we describe
may be applied over any declination range.  Future work will aim to extend this analysis
to other declination ranges, and to tie the flux calibration of these other declination ranges
back to the absolute flux scale we derive here.
